The Concurrent Calculi Formalisation Benchmark

### 2023-10-16 paper/paper.tex
just style

\subsubsection{General design remarks.}
Like the previous POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.  Still, these (and other) proof mechanisation applications need the basic techniques that our challenge problems exercise:
as per our design goal \ref{item:goal-reusability},
our problems are drawn from the basic metatheory of process calculi (without requiring the development of new theory), and focus on well-known theorems
and results that involve interesting proof techniques that may be reused in further work.
To achieve our design goal \ref{item:goal-comperison-accessibility},
we have designed our challenge problems to exercise the three aspects (linearity, scope extrusion, coinduction) independently, and they can be solved individually and in any order;
each problem should be small and easy to understand with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features we want to exercise, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing tangible
benefits and insights.\footnote{It is of course possible to reintroduce the omitted
constructs, but we leave this as a possibility for a future benchmark challenge.}
The minimality and uniformity of the calculi also allows us to target the design
goal \ref{item:goal-tutorials}, i.e.~encouraging the creation of guidelines and
tutorials for the mechanisation of concurrency theory.

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a safety theorem
for a session type system since this allows us to involve linear
reasoning without many definitions. Accordingly, we have chosen not to
include channel delegation and to use a reduction instead of a
transition system semantics.
Linearity is of course interesting in many other subfields of programming languages, but we have chosen to base our challenge on a session type system primarily due to the common use of session types in concurrency theory.
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we use a syntax where
restriction binds two names. This presentation makes the duality
of the two names obvious in the type system.
We have chosen a simple notion of well-formedness for our challenge,
and do not for instance consider deadlocked processes. While more
sophisticated notions of well-formedness are also interesting, the
proofs of these properties are more complicated and would thus make
the challenge more difficult and less focused on the linear
aspects.

@@

On the side of proof assistants, the current challenge proposes mechanising the
proofs of theorems. An interesting venue to explore would be to take advantage
of other proof assistant features. For example, proof assistants can often
produce certified code (via code extraction, or compiling and running their
definitions). Additionally, while our current challenge allow for proof
automation, a more specific challenge could be designed with the objective of
automating aspects of the proofs. Finally, a challenge could propose the
integration with other formal reasoning tools, namely model checkers. These
tools are in common use in the field --- and combining automated proofs with
proof assistants offers the potential to ease the path towards larger proofs.

>>>>

\subsubsection{General design remarks.}
Like the previous POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.  Still, these (and other) proof mechanisation applications need the basic techniques that our challenge problems exercise:
as per our design goal \ref{item:goal-reusability},
our problems are drawn from the basic metatheory of process calculi (without requiring the development of new theory), and focus on well-known theorems
and results that involve interesting proof techniques that may be reused in further work.
To achieve our design goal \ref{item:goal-comperison-accessibility},
we have drafted our challenge problems to exercise the three aspects (linearity, scope extrusion, coinduction) independently, so that they can be solved individually and in any order;
each problem should be reasonably small and easy to understand with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to exercise, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing tangible
benefits and insights.\footnote{It is of course possible to reintroduce the omitted
constructs, but we leave this as a possibility for a future benchmark challenge.}
The minimality and uniformity of the calculi also allows us to target the design
goal \ref{item:goal-tutorials}, i.e.~encouraging the creation of guidelines and
tutorials for the mechanisation of concurrency theory.

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a session type system since this allows us to involve linear
reasoning without many definitions. Accordingly, we have chosen not to
include channel delegation and to use a reduction instead of a
transition system semantics.
Linearity is of course interesting in many other subfields of programming languages, but we have chosen to base our challenge on a session type system primarily due to the common use of session types in concurrency theory.
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we use a syntax where
restriction binds two names. This presentation makes the duality
of the two names obvious in the type system.
We have chosen a simple notion of well-formedness for our challenge,
and do not for instance consider deadlocked processes. While more
sophisticated notions of well-formedness are also interesting, the
proofs of these properties are more complicated and would thus make
the challenge more difficult and less focused on the linear
aspects.

@@

On the side of proof assistants, the current challenge proposes mechanising the
proofs of theorems. An interesting venue to explore would be to take advantage
of other proof assistant features. For example, proof assistants can often
produce certified code (via code extraction, or compiling and running their
definitions). Additionally, while our current challenge is agnostic about proof
automation, a more specific challenge could be designed with the objective of
automating aspects of the proofs. Finally, a challenge could propose the
integration with other formal reasoning tools, namely model checkers. These
tools are in common use in the field --- and combining automated proofs with
proof assistants offers the potential to ease the path towards larger proofs.
